---
title: "Duke Sports Analytics Club: Technical Workshop"
author: "Jack Lichtenstein"
date: "`r Sys.Date()`"
output:
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300, cache.lazy = FALSE,
                      tidy = "styler", fig.width = 8, fig.height = 5)
```

## Loading libraries

```{r load libraries}
library(tidyverse)
theme_set(theme_light()) # setting a theme
```

## nflfastR

Let's take a look at NFL play by play data using the `nflfastR` package!

```{r}
# load package
library(nflfastR)

# load in some play-by-play data
nfl_pbp <- nflfastR::load_pbp(seasons = 2014:2020)
```

```{r}
# find offensive EPA by team for a given season
off_epa <- nfl_pbp %>% 
  filter(!is.na(posteam),
         !is.na(epa),
         !is.na(down),
         penalty == 0,
         play_type %in% c("pass", "run")) %>% 
  group_by(season, posteam) %>% 
  summarise(plays = n(),
            passes = sum(play_type == "pass"),
            rushes = sum(play_type == "run"),
            epa_pass = mean(epa[play_type == "pass"]),
            epa_run = mean(epa[play_type == "run"]))

# find defensive EPA by team for a given season
def_epa <- nfl_pbp %>% 
  filter(!is.na(posteam),
         !is.na(epa),
         !is.na(down),
         penalty == 0,
         play_type %in% c("pass", "run")) %>% 
  group_by(season, defteam) %>% 
  summarise(plays = n(),
            passes = sum(play_type == "pass"),
            rushes = sum(play_type == "run"),
            epa_pass = mean(epa[play_type == "pass"]),
            epa_run = mean(epa[play_type == "run"]))

# bind together
epa <- bind_rows(off_epa %>% 
                   mutate(type = "Offense") %>% 
                   rename(team = posteam),
                 def_epa %>%
                   mutate(type = "Defense") %>% 
                   rename(team = defteam))
```

```{r}
# how are offensive/defensive EPA rush/pass related
# are good offenses good at both rush/pass?
# are good defenses good at both rush/pass?
epa %>% 
  ggplot(aes(epa_run, epa_pass)) +
  geom_smooth(method = "lm") +
  geom_point() +
  facet_wrap(~ type) +
  labs(x = "EPA/rush",
       y = "EPA/pass")
```

Let's take a look at the stability of EPA from year to year, both on offense and defense as well as by rush and pass.

```{r}
lagged_epa <- epa %>% 
  group_by(team, type) %>% 
  # find previous year epa
  mutate(prev_epa_pass = lag(epa_pass),
         prev_epa_run = lag(epa_run)) %>% 
  ungroup()

lagged_epa %>% 
  ggplot(aes(prev_epa_pass, epa_pass)) +
  geom_smooth(method = "lm") +
  geom_point() +
  facet_wrap(~ type) +
  labs(x = "Previous year EPA/pass",
       y = "Current year EPA/pass")

# 0.138 r-squared
# pass offense from year-to-year is stable!
lagged_epa %>% 
  filter(type == "Offense") %>% 
  lm(epa_pass ~ prev_epa_pass, data = .) %>% 
  broom::glance() %>% 
  pull(r.squared)

# look at stability of all metrics
lagged_epa %>% 
  select(-c(plays:rushes)) %>% 
  pivot_longer(cols = c(epa_pass, epa_run),
               names_to = "current",
               names_prefix = "epa_") %>% 
  pivot_longer(cols = c(prev_epa_pass, prev_epa_run),
               names_to = "prev",
               values_to = "prev_value",
               names_prefix = "prev_epa_") %>% 
  # filter(name == prev) %>% 
  # filter(name == "pass") %>% 
  # lm(value ~ prev_value, data = .) %>% 
  # broom::glance()
  nest(data = c(season, team, value, prev_value)) %>% 
  mutate(lm = map(data, ~ lm(value ~ prev_value, data = .)),
         glanced = map(lm, broom::glance)) %>% 
  hoist(glanced, r_2 = "r.squared",
        p_value = "p.value") %>% 
  transmute(type, target = current,
            predictor = paste0("Previous year ", prev),
            r_2, p_value) %>% 
  mutate(across(target:predictor, ~ paste0(., " EPA"))) %>% 
  arrange(desc(r_2)) %>% 
  kableExtra::kable(format = "markdown")
```

Passing offense is most stable year to year! This is followed by pass defense

## Web scraping

Let's try some web-scraping! We will look to scrape NFL draft big boards for the upcoming 2021 NFL Draft.

- [CBS](https://www.cbssports.com/nfl/draft/prospect-rankings/)
- [Walter Football](https://walterfootball.com/nfldraftbigboard/)
- [Drafttek](https://www.drafttek.com/2021-NFL-Draft-Big-Board/Top-NFL-Draft-Prospects-2021-Page-1.asp)

```{r}
# load library
library(rvest)

# CBS
url <- "https://www.cbssports.com/nfl/draft/prospect-rankings/"

html <- read_html(url)

# html_table is great
cbs <- html %>% 
  html_table() %>% 
  pluck(1) %>% 
  janitor::clean_names() %>% 
  filter(row_number() %% 2 == 1) %>% 
  select(-c(x:x_2)) %>% 
  mutate(across(c(rk, pos_rk, wt),
                as.numeric)) %>% 
  rename(ovr_rk = rk,
         player_name = player)
```

```{r}
# Walter football
url <- "https://walterfootball.com/nfldraftbigboard/"

html <- read_html(url)

# pull out the text
text <- html %>% 
  html_elements(".divPlayerRanking") %>% 
  html_text() %>% 
  str_squish()

# put text in tibble and extract information
wf <- tibble(text = text) %>% 
  mutate(ovr_rk = str_extract(text, "\\d{1,3}\\."),
         ovr_rk = readr::parse_number(ovr_rk)) %>% 
  filter(!is.na(ovr_rk)) %>% 
  tidyr::separate(text,
                  into = c("player_name",
                           "pos_abbr",
                           "rest"),
                  sep = ", ",
                  extra = "merge") %>% 
  mutate(player_name = str_remove(player_name, "\\d{1,3}\\. "))
```

Now that we have two big boards, let's compare!

```{r}
consensus_bb <- bind_rows(cbs %>% mutate(type = "CBS"),
                          wf %>% mutate(type = "WF"))

# join in to compare ranks
joined_bb <- cbs %>% 
  left_join(wf,
            by = "player_name",
            suffix = c("_cbs", "_wf"))

# how do these big boards compare?
joined_bb %>% 
  ggplot(aes(ovr_rk_cbs, ovr_rk_wf)) +
  geom_text(aes(label = player_name),
            check_overlap = T, size = 3,
            hjust = 1, vjust = 1) +
  geom_point() +
  scale_x_reverse() +
  scale_y_reverse() +
  labs(x = "CBS big board rank",
       y = "Walter Football big board rank")

# who has the largest difference in rank?
joined_bb %>% 
  mutate(diff_rk = abs(ovr_rk_cbs - ovr_rk_wf)) %>% 
  arrange(desc(diff_rk)) %>% 
  head(20) %>% 
  select(player_name, ovr_rk_cbs, ovr_rk_wf, diff_rk) %>% 
  mutate(player_name = fct_reorder(player_name, diff_rk)) %>% 
  ggplot(aes(diff_rk, player_name)) +
  geom_col() +
  labs(x = "Difference in big board rank between CBS and Walter Football",
       y = NULL)
```

Jared Goldwire has an enormous difference in big board ranks!

## College basketball data sources

Let's explore a little bit of college basketball!

```{r}
# ncaahoopR and kenpomR are both wrappers for ESPN
# play-by-play
library(ncaahoopR)
library(kenpomR)

# these didn't work for some reason
# game_ids <- ncaahoopR::get_game_ids("Duke", season = 2020)
# espn_schedule <- kenpomR::cbb_espn_scoreboard(2020)

# play-by-play of Duke vs. Louisville
espn_pbp <- kenpomR::cbb_espn_game_all("401300969")

head(espn_pbp$Plays)

head(espn_pbp$Player)

```

How about `gamezoneR`?

```{r}
library(gamezoneR)

schedule <- gamezoneR::get_team_schedule("Duke")

pbp_duke <- gamezoneR::gamezone_cbb_pbp(schedule$game_id[30])

gamezoneR::base_court +
  geom_point(data = pbp_duke,
             aes(loc_x, loc_y,
                 color = shot_outcome))
```


